---
title: "Nice hillshaded maps in R"
output:
  md_document:
    variant: markdown_github
date: "2025-09-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

The first step is to download the needed topographic data using the `elevatr` pacakge. For that, I need to define a bounding box for my map. As an example here, I am using the coordinates of the archaeological site of Mukila, in the Kwango province of the Democratic Republic of the Congo. If you want to know more about the site, than please check out our paper here <https://www.sciencedirect.com/science/article/pii/S0277379124002531>.

```{r}
bbox <- data.frame(
  x = c(16.9862 - .01, 
        16.9862 + .01),
  y = c(-5.0156 - .01, 
        -5.0156 + .01))
```

Now I can download the SRMT background data:

```{r}
dem <- elevatr::get_elev_raster(
  locations = bbox, 
  prj = "EPSG:4326", 
  z = 14, 
  clip = "bbox")
```

The resulting raster can easily inspected using the `raster::plot()` function:

```{r}
raster::plot(dem)
```

In the next step, I create the hillshade by calculating the slope and aspect for each raster cell using the `raster::terrain()` function. The hillshade is calculated based on slope and aspect using the `raster::hillshade()` function.

```{r}
slope = raster::terrain(dem, opt = 'slope')
aspect = raster::terrain(dem, opt = 'aspect')
hill = raster::hillShade(slope, aspect, 40, 270)
```

Okay, now that we have all the necessary data, I convert them to dataframes to be able to better work with them:

```{r}
dem.df <- as(dem, "SpatialPixelsDataFrame")
dem.df <- as.data.frame(dem.df)
colnames(dem.df) <- c("value", "x", "y")

head(dem.df)
```

Now I need to get the color values for the choose color pallet. For this, I use the `colourvalues` package:

```{r}
dem.df.rgb <- cbind(
  dem.df,
  # hex values
  data.frame(
    hex = colourvalues::color_values(dem.df$value, "terrain")
  ),
  # rgb values
  colourvalues::color_values_rgb(
    dem.df$value, 
    "terrain", 
    include_alpha = F) %>% 
    as.data.frame() %>%
    dplyr::rename(
      R = V1, 
      G = V2, 
      B = V3) %>%
    dplyr::mutate(
      R = R / 255, 
      G = G / 255, 
      B = B / 255
    )
)
head(dem.df.rgb)
```

At this stage it is important to devide the color values by 255. Otherwise, the 'multiplication' later would not work.

```{r}
cowplot::plot_grid(
  # version with terrain colors applied to values:
  ggplot() + 
    geom_tile(
      data = dem.df.rgb, 
      aes(x = x, y = y, fill = value)
    ) + 
    scale_fill_gradientn(colours = terrain.colors(7)) + 
    scale_x_continuous(
      expand = c(0, 0)) + 
    scale_y_continuous(
      expand = c(0, 0)) + 
    coord_equal() + 
    ggthemes::theme_base() + 
    theme(
      plot.background = element_rect(colour = NA),
      axis.title = element_blank(), 
      legend.position = c(.01, .01), 
      legend.justification = c(0, 0),
      legend.title = element_blank()
    ),
  # version with the hex color values from above
  ggplot() + 
    geom_tile(
      data = dem.df.rgb, 
      aes(x = x, y = y, fill = hex)
    ) + 
    scale_fill_identity() + 
    scale_x_continuous(
      expand = c(0, 0)) + 
    scale_y_continuous(
      expand = c(0, 0)) + 
    coord_equal() + 
    ggthemes::theme_base() + 
    theme(
      plot.background = element_rect(colour = NA),
      axis.title = element_blank()
    ),
  nrow = 1
)
```

The `colourvalues` package seems to nicely represent the elevation, although with a slightly different spread of the colors towards the min and max.

But for this exercise, I go further and get RGB (greyscale) color values for the hillshade. But first, I convert the raster to a dataframe:

```{r}
hill.df <- as(hill, "SpatialPixelsDataFrame")
hill.df <- as.data.frame(hill.df)
colnames(hill.df) <- c("value", "x", "y")
head(hill.df)
```

Now for the RGB values, I chose a different approach, as I am only dealing with a greyscales:

```{r}
hill.df.rgb <- hill.df %>%
  dplyr::filter(!is.na(value)) %>%
  dplyr::mutate(
    R = (value - min(value)) / (max(value) - min(value)), # * 1.6 
    G = (value - min(value)) / (max(value) - min(value)), # * 1.6
    B = (value - min(value)) / (max(value) - min(value))) # * 1.6

hill.df.rgb$R[hill.df.rgb$R > 1] <- 1 
hill.df.rgb$G[hill.df.rgb$G > 1] <- 1 
hill.df.rgb$B[hill.df.rgb$B > 1] <- 1 

head(hill.df.rgb)
```

With `dem.df.rgb` and `hill.df.rgb`, we can simply multiply the values for each cell and, in case a value goes beyong 255, cap the scale at 255 for each chanel.

```{r}
dem.hill.rgb.df <- dem.df.rgb %>%
  dplyr::left_join(hill.df.rgb, by = c("x", "y")) %>%
  dplyr::mutate(R = R.x * R.y,
                G = G.x * G.y,
                B = B.x * B.y) %>%
  dplyr::filter(!is.na(R))

dem.hill.rgb.df$col <- rgb(
  dem.hill.rgb.df$R * 255, 
  dem.hill.rgb.df$G * 255, 
  dem.hill.rgb.df$B * 255, 
  maxColorValue = 255)

head(dem.hill.rgb.df)
```

```{r}
ggplot() + 
  geom_tile(
    data = dem.hill.rgb.df, 
    aes(x = x, y = y, fill = col)
  ) + 
  scale_fill_identity() + 
  scale_x_continuous(
    expand = c(0, 0)) + 
  scale_y_continuous(
    expand = c(0, 0)) + 
  coord_equal() + 
  ggthemes::theme_base() + 
  theme(
    plot.background = element_rect(colour = NA),
    axis.title = element_blank()
  )
```

This looks like an interesting hillshaded map :)